\bf Effects without a Cause: The Search for Demographic Anomalies

### 2020-01-14 MANUSCRIPT/manuscript.Rmd
Working through figures and associated text for the Main Document.

library(knitr)
read_chunk('../R/SCRIPTS/000-Libraries.R')
read_chunk('../R/SCRIPTS/001-DataLoad.R')
```

@@

@@

Abductive modeling is the movement from the inductive to the deductive, and sometimes back and forth, to reach conclusions [@bryant2014realm] or “inferring cause from effect” [@Crowder2017]. In situations with large data sets, an abductive approach is far superior to deductive hypothesis testing, as p-values with an extremely large number of cases are far from revealing [@head2015extent;@nuzzo2014scientific] and we do not want to make purely inductive inferences from data of questionable generalizability [@ruggles2014big]. Owing to the long history of big datasets in demographic research (one of the original "big data" resources [@ruggles2014big]), the rich demographic data available in the United States make the potential revelation of interesting and important demographic phenomena not only possible, but extremely plausible – even if identification of the phenomena occurs without identifying the underlying cause. In other sciences, the identification of effects without causes has led to new theories of galactic migration of planets and other breakthroughs [@gomes2005n].

@@

where $\pi(B)=\sum_{i=o}^{inf} \pi_iB^i$. The identification of outliers then involves a three step process to (1) identify all potential outliers, $t_j$ and $L_j (B)$, (2) joint estimates of model parameters and outlier effects are computed to identify potentially spurious outliers, and (3) the outliers and effects are re-estimated without spurious outliers. Of importance here is step 2, which relies on some critical value above which an outlier at time point m is considered spurious. Based on Chien and Liu’s [-@chen1993joint] recommendation, we set a critical value of 3.5 which generally minimizes the possibility of Type I errors or false positive outliers. Outliers are reported using a simple t-statistic.

@@

# df2<- setNames(data.frame(matrix(ncol = 5, nrow = 0)), c("type", "ind", "time", "coefhat", "tstat"))
dat2 <- filter(dat_mort, location == "X36")
dat3 <- ts(dat2$Target, start = year(min(dat_mort$Year)), end = c(year(max(dat_mort$Year)),12), frequency = 12)
outlier.county <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)

plot(outlier.county,
     cex.axis=0.8) +
title(main = "New York - Monthly Mortality",
      adj = 1)
```

@@

# df2<- setNames(data.frame(matrix(ncol = 5, nrow = 0)), c("type", "ind", "time", "coefhat", "tstat"))
dat2 <- filter(dat_mort, location == "X33")
dat3 <- ts(dat2$Target, start = year(min(dat_mort$Year)), end = c(year(max(dat_mort$Year)),12), frequency = 12)
outlier.county <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)

plot(outlier.county,
     cex.axis=0.8) +
title(main = "New Hampshire - Monthly Mortality",
      adj = 1)
```

@@

>>>>

library(knitr)
read_chunk('../R/SCRIPTS/000-Libraries.R')
read_chunk('../R/SCRIPTS/001-DataLoad.R')
read_chunk('../R/SCRIPTS/002-StatesAnalysis.R')
```

@@

```{r StatesAnalysis, include=FALSE}
```

@@

Abductive modeling is the movement from the inductive to the deductive, and sometimes back and forth, to reach conclusions [@bryant2014realm] or “inferring cause from effect” [@Crowder2017]. In situations with large data sets, an abductive approach is far superior to deductive hypothesis testing, as p-values with an extremely large number of cases are far from revealing [@head2015extent;@nuzzo2014scientific] and we do not want to make purely inductive inferences from data of questionable generalizability [@ruggles2014big]. Owing to the long history of big data sets in demographic research (one of the original "big data" resources [@ruggles2014big]), the rich demographic data available in the United States make the potential revelation of interesting and important demographic phenomena not only possible, but extremely plausible – even if identification of the phenomena occurs without identifying the underlying cause. In other sciences, the identification of effects without causes has led to new theories of galactic migration of planets and other breakthroughs [@gomes2005n].

@@

where $\pi(B)=\sum_{i=o}^{inf} \pi_iB^i$. The identification of outliers then involves a three step process to (1) identify all potential outliers, $t_j$ and $L_j (B)$, (2) joint estimates of model parameters and outlier effects are computed to identify potentially spurious outliers, and (3) the outliers and effects are re-estimated without spurious outliers. Of importance here is step 2, which relies on some critical value above which an outlier at time point m is considered spurious. Based on Chen and Liu’s [-@chen1993joint] recommendation, we set a critical value of 3.5 which generally minimizes the possibility of Type I errors or false positive outliers. Outliers are reported using a simple t-statistic.

@@

plot(outlier.mort.ny,
     cex.axis=0.8) +
title(main = "New York - Monthly Mortality",
      adj = 1)
```

@@

plot(outlier.mort.nh,
     cex.axis=0.8) +
title(main = "New Hampshire - Monthly Mortality",
      adj = 1)
```

@@

## Fertility

We begin with a clear temporary change (TC) outlier for fertility in Louisiana (**\autoref{fig:fertla}**). Here we detect two TC outliers, nearly back-to-back in 2005 (2005:08 *t* = `r prettyNum(outlier.fert.la$outliers$tstat[1], digits = 3)`; and 2005:10 *t* = `r prettyNum(outlier.fert.la$outliers$tstat[2], digits = 3)`), coinciding with the destruction of Hurricane Katrina from the same year. The migration associated with Hurricane Katrina has garnered most of the attention of demographers [@fussellRecoveryMigrationCity2014;@horiDisplacementDynamicsSouthern2009], but the hurricane had a very clear impact on fertility behaviors too, creating a mini "baby bust" in Louisiana. We estimate `r prettyNum(abs(sum(outlier.fert.la$y - outlier.fert.la$yadj)), digits = 0, big.mark = ",")` fewer births in Louisiana compared to the counterfactual, likely attributable to the Hurricane.

```{r FertilityLouisiana, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Louisiana state fertility, 2003-2018.} Here we detect two outliers, back-to-back, likely resulting from Hurricane Katrina August and October 2005, representing a decrease of more than 4,400 births due to the hurricane. \\label{fig:fertla}")}

plot(outlier.fert.la,
     cex.axis=0.8) +
  title(main = "Louisiana - Monthly Fertility",
        adj = 1)
```

We also highlight a level shift (LS) outlier for fertility in Connecticut. Here we detect a strong (*t*= `prettyNum(outlier.fert.conn$outliers$tstat, digits = 3)`; `r prettyNum(outlier.fert.conn$outliers$coefhat, digits = 0)` births/month) level shift toward lower fertility beginning in August 2009 (2009:08) that continues through the end of the period (**\autoref{fig:fertconn}**). This LS toward lower fertility reduced the # of births by `r prettyNum(sum(outlier.fert.conn$y - outlier.fert.conn$yadj), digits = 0, big.mark = ",")` or `r percent(sum(outlier.fert.conn$y - outlier.fert.conn$yadj) / sum(outlier.fert.conn$yadj[80:180]))` lower than the counterfactual time series. The stock market crash of 2008, where the Dow Jones Industrial Average had the largest single-day loss up to that point, occured just 11 months before we detect a shift toward lower fertility. We believe the trend toward lower fertility in August 2009 and the stock market crash in September 2008 are linked.

```{r FertilityConnecticut, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Hawaiian state fertility, 2003-2018.} Here we detect one outlier, an additive outlier (AO) in May 2014 ( \textit{t}=4.96), representing a 14% increase in expected births in that month. \\label{fig:fertconn}")}

plot(outlier.fert.conn,
     cex.axis=0.8) +
  title(main = "Connecticut - Monthly Fertility",
        adj = 1)
```

## Interesting Anomalous Fertility/Mortality events
In the examples above, we highlighted four fertility/mortality anomalies with plausible explanations. However, these four anomalies are but a fraction of anomalies we detect. Here, we highlight four additional anomalies without easily identifiable explanations.

In **\autoref{fig:ferthawaii}** we identify a single additive anomaly in Hawaiian fertility in May 2014. This is a strong anomaly with a *t*-statistic of 4.96, 14% above the counter-factual time series. This single, anomalous month is also the second highest monthly births in the time series. We have no plausible explanation for this anomaly. We do not believe this is simply a data error as the other extreme values, September 2008 and February 2005 with the highest and lowest recorded fertility respectively, were not identified as anomalous events. Even if we were to assume this anomaly resulted from data entry error, it remains an *unaltered* data error in the Hawaiian monthly fertility data.

```{r AnomalyHawaii, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Hawaiian state fertility, 2003-2018.} Here we detect one outlier, an additive outlier (AO) in May 2014 ( \textit{t}=4.96), representing a 14% increase in expected births in that month. \\label{fig:ferthawaii}")}

plot(outlier.fert.hi,
     cex.axis=0.8) +
  title(main = "Hawaii - Monthly Fertility",
        adj = 1)
```

We detect a very interesting, unexplainable mortality anomaly in North Carolina (**\autoref{fig:fertmiss}**). Beginning in February 2008 , North Carolina saw a level shift (LS) (2008:02; *t*-stat = `r prettyNum(outlier.mort.nc$outliers$tstat[3], digits = 3)`) of 800 fewer deaths per month until the end of the period. Granted, the decade after the LS still saw three mortality anomalies that resulted in additional, unexpected deaths in 2009, 2012, and 2017, the LS is still a remarkable mortality reduction of nearly 100k (`r prettyNum(sum(outlier.mort.nc$y[62:180] - outlier.mort.nc$yadj[62:180]), digits = 0, big.mark = ",")`) deaths over 10 years! While North Carolina saw the raw count of mortality remain constant or slightly reduce since 2008, the state grew from 9.3 million residents in 2008 to 10.4 million in 2018. In spite of its growing population, North Carolina has, rather miraculously, *reduced* its overall mortality risk.
```{r AnomalyMississippi, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in North Carolina state mortality, 1999-2018.} Here we detect one outlier, an additive outlier (AO) in May 2014 ( \textit{t}=4.96), representing a 14% increase in expected births in that month. \\label{fig:fertmiss}")}

plot(outlier.mort.nc,
     cex.axis=0.8) +
  title(main = "North Carolina - Monthly Mortality",
        adj = 1)
```

Hawaii - AO
X33 - LS
X28
X22
X20
X05

### 2020-01-14 MANUSCRIPT/manuscript.tex
Working through figures and associated text for the Main Document.

Abductive modeling is the movement from the inductive to the deductive,
and sometimes back and forth, to reach conclusions
\citep{bryant2014realm} or ``inferring cause from effect''
\citep{Crowder2017}. In situations with large data sets, an abductive
approach is far superior to deductive hypothesis testing, as p-values
with an extremely large number of cases are far from revealing
\citep{head2015extent, nuzzo2014scientific} and we do not want to make
purely inductive inferences from data of questionable generalizability
\citep{ruggles2014big}. Owing to the long history of big datasets in
demographic research (one of the original ``big data'' resources
\citep{ruggles2014big}), the rich demographic data available in the
United States make the potential revelation of interesting and important
demographic phenomena not only possible, but extremely plausible -- even
if identification of the phenomena occurs without identifying the
underlying cause. In other sciences, the identification of effects
without causes has led to new theories of galactic migration of planets
and other breakthroughs \citep{gomes2005n}.

@@

where \(\pi(B)=\sum_{i=o}^{inf} \pi_iB^i\). The identification of
outliers then involves a three step process to (1) identify all
potential outliers, \(t_j\) and \(L_j (B)\), (2) joint estimates of
model parameters and outlier effects are computed to identify
potentially spurious outliers, and (3) the outliers and effects are
re-estimated without spurious outliers. Of importance here is step 2,
which relies on some critical value above which an outlier at time point
m is considered spurious. Based on Chien and Liu's
\citeyearpar{chen1993joint} recommendation, we set a critical value of
3.5 which generally minimizes the possibility of Type I errors or false
positive outliers. Outliers are reported using a simple t-statistic.

@@

>>>>

Abductive modeling is the movement from the inductive to the deductive,
and sometimes back and forth, to reach conclusions
\citep{bryant2014realm} or ``inferring cause from effect''
\citep{Crowder2017}. In situations with large data sets, an abductive
approach is far superior to deductive hypothesis testing, as p-values
with an extremely large number of cases are far from revealing
\citep{head2015extent, nuzzo2014scientific} and we do not want to make
purely inductive inferences from data of questionable generalizability
\citep{ruggles2014big}. Owing to the long history of big data sets in
demographic research (one of the original ``big data'' resources
\citep{ruggles2014big}), the rich demographic data available in the
United States make the potential revelation of interesting and important
demographic phenomena not only possible, but extremely plausible -- even
if identification of the phenomena occurs without identifying the
underlying cause. In other sciences, the identification of effects
without causes has led to new theories of galactic migration of planets
and other breakthroughs \citep{gomes2005n}.

@@

where \(\pi(B)=\sum_{i=o}^{inf} \pi_iB^i\). The identification of
outliers then involves a three step process to (1) identify all
potential outliers, \(t_j\) and \(L_j (B)\), (2) joint estimates of
model parameters and outlier effects are computed to identify
potentially spurious outliers, and (3) the outliers and effects are
re-estimated without spurious outliers. Of importance here is step 2,
which relies on some critical value above which an outlier at time point
m is considered spurious. Based on Chen and Liu's
\citeyearpar{chen1993joint} recommendation, we set a critical value of
3.5 which generally minimizes the possibility of Type I errors or false
positive outliers. Outliers are reported using a simple t-statistic.

@@

\hypertarget{fertility}{%
\subsection{Fertility}\label{fertility}}

We begin with a clear temporary change (TC) outlier for fertility in
Louisiana (\textbf{\autoref{fig:fertla}}). Here we detect two TC
outliers, nearly back-to-back in 2005 (2005:08 \emph{t} = -6; and
2005:10 \emph{t} = -5), coinciding with the destruction of Hurricane
Katrina from the same year. The migration associated with Hurricane
Katrina has garnered most of the attention of demographers
\citep{fussellRecoveryMigrationCity2014, horiDisplacementDynamicsSouthern2009},
but the hurricane had a very clear impact on fertility behaviors too,
creating a mini ``baby bust'' in Louisiana. We estimate 4,411 fewer
births in Louisiana compared to the counterfactual, likely attributable
to the Hurricane.

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/FertilityLouisiana-1.pdf}
\caption{\textbf{Anomaly detection in Louisiana state fertility, 2003-2018.}
Here we detect two outliers, back-to-back, likely resulting from
Hurricane Katrina August and October 2005, representing a decrease of
more than 4,400 births due to the hurricane. \label{fig:fertla}}
\end{figure}

We also highlight a level shift (LS) outlier for fertility in
Connecticut. Here we detect a strong (\emph{t}=
\texttt{prettyNum(outlier.fert.conn\$outliers\$tstat,\ digits\ =\ 3)};
-285 births/month) level shift toward lower fertility beginning in
August 2009 (2009:08) that continues through the end of the period
(\textbf{\autoref{fig:fertconn}}). This LS toward lower fertility
reduced the \# of births by -28,752 or -8.57\% lower than the
counterfactual time series. The stock market crash of 2008, where the
Dow Jones Industrial Average had the largest single-day loss up to that
point, occured just 11 months before we detect a shift toward lower
fertility. We believe the trend toward lower fertility in August 2009
and the stock market crash in September 2008 are linked.

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/FertilityConnecticut-1.pdf}
\caption{\textbf{Anomaly detection in Hawaiian state fertility, 2003-2018.}
Here we detect one outlier, an additive outlier (AO) in May 2014 (
extit\{t\}=4.96), representing a 14\% increase in expected births in
that month. \label{fig:fertconn}}
\end{figure}

\hypertarget{interesting-anomalous-fertilitymortality-events}{%
\subsection{Interesting Anomalous Fertility/Mortality
events}\label{interesting-anomalous-fertilitymortality-events}}

In the examples above, we highlighted four fertility/mortality anomalies
with plausible explanations. However, these four anomalies are but a
fraction of anomalies we detect. Here, we highlight four additional
anomalies without easily identifiable explanations.

In \textbf{\autoref{fig:ferthawaii}} we identify a single additive
anomaly in Hawaiian fertility in May 2014. This is a strong anomaly with
a \emph{t}-statistic of 4.96, 14\% above the counter-factual time
series. This single, anomalous month is also the second highest monthly
births in the time series. We have no plausible explanation for this
anomaly. We do not believe this is simply a data error as the other
extreme values, September 2008 and February 2005 with the highest and
lowest recorded fertility respectively, were not identified as anomalous
events. Even if we were to assume this anomaly resulted from data entry
error, it remains an \emph{unaltered} data error in the Hawaiian monthly
fertility data.

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/AnomalyHawaii-1.pdf}
\caption{\textbf{Anomaly detection in Hawaiian state fertility, 2003-2018.}
Here we detect one outlier, an additive outlier (AO) in May 2014 (
extit\{t\}=4.96), representing a 14\% increase in expected births in
that month. \label{fig:ferthawaii}}
\end{figure}

We detect a very interesting, unexplainable mortality anomaly in North
Carolina (\textbf{\autoref{fig:fertmiss}}). Beginning in February 2008 ,
North Carolina saw a level shift (LS) (2008:02; \emph{t}-stat = -4.76)
of 800 fewer deaths per month until the end of the period. Granted, the
decade after the LS still saw three mortality anomalies that resulted in
additional, unexpected deaths in 2009, 2012, and 2017, the LS is still a
remarkable mortality reduction of nearly 100k (-88,335) deaths over 10
years! While North Carolina saw the raw count of mortality remain
constant or slightly reduce since 2008, the state grew from 9.3 million
residents in 2008 to 10.4 million in 2018. In spite of its growing
population, North Carolina has, rather miraculously, \emph{reduced} its
overall mortality risk.
\includegraphics{manuscript_files/figure-latex/AnomalyMississippi-1.pdf}

Hawaii - AO X33 - LS X28 X22 X20 X05
